Left-Linear Rewriting in Adhesive Categories

### 2024-07-08 arxiv_concur/main.tex
arxiv

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.  This is an apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

%\tableofcontents

\iffalse
\section*{Summary (with dependencies)}

\subsection*{REWRITING}
\begin{itemize}
\item DPO Rewriting in $\mathcal{M}$-adhesive categories, with left-$\mathcal{M}$-linear rules (from now on shortened in ``linear'')\\
  Observe that rewriting is deterministic (by uniqueness of
  PO-complement Corollary~\ref{lem:radj})

\item Category of abstract derivation for rewriting system $(\X, \R)$
  \begin{itemize}
  \item \emph{DPO-derivation category} $\dpo$: objects are graphs,
    arrows between possibly empty, derivations.
  \item \emph{DPO-abstract decorated derivation category} $\dpi$
  \end{itemize}

\item \emph{Consistent permutations} (Definition~\ref{def:permcon}):
  permutation on the steps of two abstract decorated derivations and
  isomorphism between the colimits such that arrows into the colimits
  ``commutes''.
  \begin{enumerate}
  \item It is unique for consuming
    rewriting systems (Corollary~\ref{cor:unique})
  \item They can be composed
    (Lemma~\ref{lem:sum})
  \end{enumerate}
\item

\item Notions of independence\\ We have several
  possible notions of independence for two direct derivations $\dder{D}
  ; \dder{D'}$
  \begin{enumerate}

  \item \emph{Existence of a filler} There
is a filler for $\dder{D} ; \dder{D'}$ (Definition~\ref{def:filler})

\item \emph{Switchable} There is
  $\dder{D}_1' ; \dder{D}_1$ which is a \emph{switch} of $\dder{D} ;
  \dder{D'}$ (Definition~\ref{de:switch}, being a switch is defined
  axiomatically) [actually, Definition~\ref{de:good-switchable} provides
  a definition of switchable which looks different to me, I would avoid
  this, to be checked]

\item \emph{(Weakly) Sequential
    independent} There is an \emph{independence pair}
  (Definition~\ref{de:sequential-independence}) They are called
  \emph{strongly sequential independent} if the independence pair is
  unique.
\end{enumerate}

We have that $(1) \Rightarrow (2)$ (Theorem~\ref{prop:fil}) and
$(2) \Rightarrow (3)$ while we expect that in general
$(3) \not\Rightarrow (1)$ (but no counterexample). This is instead
known to hold for linear rules.

Some terminology (Definition~\ref{de:good-switchable})
\begin{itemize}

\item \emph{Strong Enforcing rewriting systems}\\ An
  independence pairs induced by a filler is called \emph{strong} A
  rewriting system where all independence pairs are strong is called
  \emph{strong enforcing}

  NOTE: the categories in which
  condition $B$ from~\cite{baldan2011adhesivity} identifies a large
  class of (quasi)adhesive categories where left-linear systems are all
  strong enforcing. Adapted here to M-adhesivity.

\item \emph{Well-Switching rewriting
    systems}\\ Strong Enforcing and existing independence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

\item \emph{Switch equivalent derivations}
  (concrete)\\
  Definition~\ref{de:switchable}

  THEOREM: Two switch equivalent derivations admit a consistent
  permutation (Lemma~\ref{lem:consperm} ??).

  A consequence of this used for proving that the slice of traces is a
  pre-order.

  The converse holds only when rules are linear [proof clear, but not
  written]  
\end{itemize}

\subsection*{CONCURRENCY}

This part is dealt with for well-switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well-switching
\item Left-linear rewriting systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well-switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well-switching
\end{itemize}

We show
\begin{itemize}

\item Three-steps lemma (Lemma~\ref{lem:indep-global-left}), which is a weakened version of ``independence is global''.
  \begin{itemize}

  \item If in $abc$ I have that $ab$ are independent and $c$ is anticipated, they remain independent.
  \item  If in $abc$ I have that $bc$ are independent and $a$ is moved to the end, they remain independent only if there is an additional condition (essentially requiring that $a$ was not a cause of $c$)
  \end{itemize}

\item No need of useless switches (if two derivations are switch equivalent I can obtain one from another applying only inversions)

\item weak prime domain e connected es semantics
\end{itemize}
\fi

@@

%\todo{A VERY NICE INTRODUCTION}
%
%SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e.~where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are the
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

A noteworthy advantage of DPO rewriting is that it can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}, which include, among others, 
toposes~\cite{johnstone2007quasitoposes} and string
diagrams~\cite{bonchi2022string}.
%
Operating within the framework of adhesive categories allows one to
devise the foundations of a rewriting theory that can be then
instantiated to the context of interest, thus avoiding the need of
repeatedly proving similar ad-hoc results for each specific
setting. % In fact, the framework is so general that it allowed to
% recast in the DPO approach many additional formalisms, ranging from
% toposes~\cite{johnstone2007quasitoposes} to string
% diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focused on %what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanism
is at the core of e-graphs (equality graphs), used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
merging nodes corresponding to equal expressions, the idea 
-- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e.~the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity}, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored,
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

% Just to get some basic intuition about left-linear rewriting systems, their capabilities and the problems which can arise, assume we want to model a situation in which a coffee comes with a bag of sugar, a portion of milk and bag of cocoa. The situation is represented by the graph $G_0$ in Fig.~\ref{fi:coffee-rewriting}: the loop $\mathit{cf}$ is the coffee and the three edges $m$, $s$ and $c$  represent the milk, sugar and cocoa, respectively. Adding one of these elements, call it $x$, in the coffee is modeled as merging the corresponding node with the coffee node so that a self-loop $x$ is created.

% The available rules are in Fig.~\ref{fi:coffee-rules}. Rules $\rho_m$ models the addition of milk and sugar, while $\rho_c$ adds cocoa and sugar to the coffee. Once the sugar has been added, the coffee can be stirred and then it is ready to drink. This is represented by rule $\rho_r$ which check for the presence of sugar in the coffee, and stirs it.

% A possible rewriting sequence is depicted in Fig.~\ref{fi:coffee-rewriting}, applying $\rho_m$, $\rho_c$ and $\rho_r$ in sequence, thus producing a coffee with milk, sugar and cocoa, ready for drinking. A first interesting observation is that after applying rule $\rho_m$ to $G_0$, thus producing graph $G_1$, we can still apply rule $\rho_c$, with a non-injective match. Intuitively, the two rules do not interfere. Rather, $\rho_c$ applied after $\rho_m$ will have no effect on the sugar edge (intuitively, because sugar has been already added). Indeed, the rules could be also be applied in reverse order. Should they be considered independent? Moreover, since rule $\rho_r$ just require the presence of sugar, it could be anticipated, and applied after $r_m$. This might lead to think that $\rho_c$ and $\rho_m$ are independent. However, this is not completely clear, since in absence of $\rho_m$, the application of $\rho_c$ would be essential to enable $\rho_r$. ``There is great chaos under heaven, [hence] the situation is excellent!''

Just to get some basic intuition about left-linear rewriting systems, their capabilities and the problems that can arise, assume we want to model a situation in which a coffee comes with a bag of white sugar and a bag of brown sugar. The situation is represented by the graph $G_0$ in Fig.~\ref{fi:coffee-rewriting}: the loop $\mathit{c}$ is the coffee, the two loops $b$ and $w$  represent the white and brown sugar. The corresponding nodes are connected to the coffee node by an $s$-labelled edge to indicate that they can be possibly added to the coffee. The available rules are in Fig.~\ref{fi:coffee-rules}. Rules $\rho_w$ and $\rho_b$ model the addition of white and brown sugar, respectively, to the coffee.
%
Note that adding one kind of sugar, say $x$, to the coffee is realised
by deleting loop $x$ and merging the corresponding node with the
coffee node so that a self-loop $s$ is created.

@@

In this paper we perform a systematic exploration of these phenomena at the
% Here we perform this exploration at the
level of $\mathcal{M}$-adhesive categories,
which encompass a large family of DPO-based formalisms.
% adopted by researchers and  practitioners alike.

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only in the order of
independent steps.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we prove that linear rewriting systems 
%on $\mathcal{M}$-adhesive categories 
are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, 
sometimes in
weakened form, most of the results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development faces a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
%\newpage
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

The fact that for left-linear rewriting systems the characterisation
of switch equivalence via consistent permutations fails, and thus
% : the existence of a consistent permutation is only necessary but
% not sufficient
cannot be used for proving (1) and (2), leads to asking if these
properties hold.  We show that much of the theory can be
retained: globality of independence holds in a weaker form,
conceptually linked to the fact that fusions introduce a form of
disjunctive causality, while consistency of switching holds
unchanged. Finally, we prove that a canonical form for switching
sequences can also be recovered. These results represent a solid basis for
developing a satisfactory theory of concurrency for left-linear
rules.

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

% \begin{notation}
%   We stipulate some notational conventions that will be used
%   throughout the paper.
%   \begin{itemize}
%   \item
%     Given a category $\X$ we will not distinguish notationally
%     between $\X$ and its class of objects, so that ``$X\in \X$'' means
%     that $X$ belongs to the class of objects of $\X$.

%   %\item
%     %If $1$ is a terminal object in $\X$, the unique
%    % arrow $X\to 1$ from an object $X$ will be denoted by
%     %$!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
%     %the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

%   \item $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
%     all arrows, monos and regular monos of $\X$, respectively.

%   \item Given an integer $n\in \mathbb{Z}$, $[0,n]$ denotes the
%     set of natural numbers less than or equal to $n$, assuming that
%     $[0,n]=\emptyset$ whenever $n<0$.
%   \end{itemize}
% \end{notation}

% \begin{figure}
%   \centering
%   \begin{subfigure}[b]{0.45\textwidth}
%     \centering
%     \xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}        
%     \caption{}
%     \label{fig:vk:square}
%   \end{subfigure}
%   \hfill
%   \begin{subfigure}[b]{0.45\textwidth}
%     \centering
%     \xymatrix@C=10pt@R=10pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D}
%     \caption{}
%     \label{fig:vk:cube}
%   \end{subfigure}
%   \caption{Van Kampen property.}
%   \label{fig:vk}
% \end{figure}

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. % We
% will recall it and examine some of its consequences.
In order to define it we need to introduce some terminology.
% and facts regarding subclasses of
% $\mor(\X)$.
%
%\begin{definition}
  Let  $\X$ be a category. A subclass $\mathcal{A}$ of
  $\mor(\X)$ is called %\todo{il diagramma confonde la composizione}
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
%\end{definition}

% So equipped, we can introduce the notion of \emph{$\mathcal{A}$-Van Kampen square}.

  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider
    %\todo{diagramS?} 
    the diagram\\
    %
    \parbox{10cm}{
      aside.
      % 
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
      \begin{enumerate}
      \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
      if and only if the front and right faces are pullbacks.
      \end{enumerate}}
    \parbox{2cm}{$\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }$ }\\
    %    
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

% old version
% \parbox{10cm}{
%   \begin{definition}[Van Kampen property]
%     Let $\X$ be a category and consider the diagrams aside.
%  %   
% %    \hspace{15pt}
%     Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
%     % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
%     is an \emph{$\mathcal{A}$-Van Kampen square} if
%     \begin{enumerate}
%     \item it is a pushout square;
%     \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
%     if and only if the front and right faces are pullbacks.
%     \end{enumerate}
%     Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

%     We call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square simply a \emph{Van Kampen} (\emph{stable}) square.
%   \end{definition}}
% \parbox{2cm}{$\xymatrix@C=10pt@R=10pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D \\A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}\\ \\C\ar[rr]_{g} &&D}$ }

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
(DPO) approach.
%We next recall the basic definitions and properties of this approach.

>>>>

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms. We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.  This is an apparently minimal, yet technically hard enhancement, since  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
via a suitable equivalence. Intuitively, the equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
Concerning formalisms based on the rewriting paradigm, concurrency often boils down to having a notion of independence between two consecutive steps. 
Noteworthy examples are the
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}.

A noteworthy advantage of DPO rewriting is that it can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}, which include, among others, toposes~\cite{johnstone2007quasitoposes} and string diagrams~\cite{bonchi2022string}. Operating within the framework of adhesive categories allows one to devise the foundations of a rewriting theory that can be then instantiated to the context of interest, thus avoiding the need of
repeatedly proving similar ad-hoc results for each specific setting.

So far, most of the theoretical results focused on \emph{linear rules}: intuitively, a step is required only to consume, preserve and generate entities.  However, in some situations it is also necessary to ``merge'' parts of the state.  This happens in the context of string diagrams mentioned above, but also in graphical implementations of nominal calculi where, as a result of name passing, the received name is identified with a local one at the receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanism is at the core of e-graphs (equality graphs), used for rewrite-driven compiler optimisations~\cite{WNW:egg}: rather thanmerging nodes corresponding to equal expressions, the idea -- conceptually and, as we will see, also mathematically similar -- is to maintain an
equivalence over the nodes.

Technically, to acquire the expressiveness needed for capturing fusions, i.e.~the merge of some elements in the state, requires to move from linear to left-linear rewriting rules. While left-linear rules have been considered in various instances (mainly in categories of graphs) and some results concerning the corresponding rewriting theory have been put forward~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity}, the phenomena arising when devising a theory of independence and concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories have not been systematically explored, even if an investigation in the setting of quasi-topoi, considering non-linear rules yet restricting their applicability, has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

Just to get some basic intuition about left-linear rewriting systems, their capabilities and the problems that can arise, assume we want to model a situation in which a coffee comes with a bag of white sugar and a bag of brown sugar. The situation is represented by the graph $G_0$ in Fig.~\ref{fi:coffee-rewriting}: the loop $\mathit{c}$ is the coffee, the two loops $b$ and $w$  represent the white and brown sugar. The corresponding nodes are connected to the coffee node by an $s$-labelled edge to indicate that they can be possibly added to the coffee. The available rules are in Fig.~\ref{fi:coffee-rules}. Rules $\rho_w$ and $\rho_b$ model the addition of white and brown sugar, respectively, to the coffee. Note that adding one kind of sugar, say $x$, to the coffee is realised by deleting loop $x$ and merging the corresponding node with the coffee node so that a self-loop $s$ is created.

@@

In this paper we perform a systematic exploration of these phenomena at the level of $\mathcal{M}$-adhesive categories, which encompass a large family of DPO-based formalisms.

As an initial step, we observe that for left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or plainly do not hold. Firstly, the notion of sequential independence and the Church-Rosser theorem, ensuring that sequential independent steps can be performed in reverse order, does not hold out of the box. Even worse, given two independent steps it might be possible to switch them in several
different ways. As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only in the order of
independent steps.

@@

On the one hand, they capture most categories of interest for rewriting. Indeed, we identify general classes of left-linear rewriting systems on graph-like structures that are well-switching. As a sanity check, we prove that linear rewriting systems are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can be developed for well-switching rewriting systems, recovering, sometimes in weakened form, most of the results that hold in the linear case. The development faces a main technical obstacle. In the literature on linear rewriting systems, a central result is the characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}. In turn this is the key to derive some fundamental properties for linear rewriting, namely
\begin{enumerate}
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent;
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the associated permutation.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

The fact that for left-linear rewriting systems the characterisation of switch equivalence via consistent permutations fails, and thus cannot be used for proving (1) and (2), leads to asking if these
properties hold.  We show that much of the theory can be retained: globality of independence holds in a weaker form, conceptually linked to the fact that fusions introduce a form of disjunctive causality, while consistency of switching holds unchanged. Finally, we prove that a canonical form for switching
sequences can also be recovered. These results represent a solid basis for developing a satisfactory theory of concurrency for left-linear rules.

\emph{Synopsis.} 
In \S\ref{sec:ade} we recall the basic definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal, introducing the standard notion of independence and a novel axiomatic notion of switchability, showing that the former fails to imply the latter for left-linear rewriting systems. We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems, and we prove that the classical results for linear rewriting systems are recovered for such a class. Finally, in \S\ref{conclusions} we outline directions for future work. The appendices recall basic results on adhesive categories (Appendix~\ref{app:ade}), present \emph{strong enforcing} rewriting systems, a class of systems where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}), and give proofs of the results (Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}

This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}. Given a category $\X$ we will not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers less than or equal to $n$; in particular, $[0,n]=\emptyset$ if $n<0$.

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. In order to define it we need to introduce some terminology.  Let  $\X$ be a category. A subclass $\mathcal{A}$ of
  $\mor(\X)$ is called
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}

  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the diagram\\
    \parbox{10cm}{
      aside.
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    is an \emph{$\mathcal{A}$-Van Kampen square} if
      \begin{enumerate}
      \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
      if and only if the front and right faces are pullbacks.
      \end{enumerate}}
    \parbox{2cm}{$\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }$ }\\
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
(DPO) approach.
